#Exercice: Principal Component Analysis

##By Kalvin Dobler


**Part 1: generation of a non-linear dataset


Two groups (A and B) are depicted in the plot below. In particular, one group is "envelopped" by the other.


```{r, echo = TRUE}
circle <- data.frame(radius = rep(c(0, 1), 500) + rnorm(1000, sd = 0.05),
                     phi = runif(1000, 0, 2 * pi), group = rep(c("A", "B"), 500))

circle <- transform(circle, x = radius * cos(phi), y = radius * sin(phi), z = rnorm(length(radius)))

circle$group = as.factor(circle$group)
colors = c("blue", "red")[circle$group]
data <- circle[,which(colnames(circle) == "x" |colnames(circle) == "y")]
plot(data, col = colors, xlab = "", ylab = "")
````

**Part 2: compute PCA

The result of the PCA returns the exact distribution of the original data. This is because the original data set cannot be linearly separated.


```{r, echo=TRUE, message = FALSE, warning = FALSE}
library(ggfortify)
data_pca <- prcomp(data, center = TRUE, scale. = FALSE)
autoplot(data_pca, data = circle, colour = colors)
```

**Part 3: try an alnertavie to solve the problem

In order to make the dataset linearly separable, I chose to add a third dimension with the following function: f(x, y) = x^2 + y^2.

```{r, echo = TRUE, message = FALSE, warning = FALSE}
library(scatterplot3d)

data$z <- (circle$x)^2 + (circle$y)^2

scatterplot3d(x = data$x, data$y, data$z, angle = 40,
              xlab = "X", ylab = "Y", zlab = "Z",
              color = colors, box = FALSE)
```

**Part 4: compute PCA

The two groups were able to be linearly separated with the addition of a third dimension.
In this case, the first two PCs explain for 70% of the total variance. In contrast, The PCA performed on the original data was not able to linearly separate it and thus, the first two PCs explained for 100% of the total variance. 

```{r, echo = TRUE, message = FALSE, warning = FALSE}
data_pca_3d = prcomp(data, center = TRUE, scale. = FALSE)
autoplot(data_pca_3d, data = circle, colour = colors)
```







